Predicting tree failure likelihood for utility risk mitigation via a novel convolutional neural network

### 2021-02-26 docs/manuscript/manuscript.tex

\begin{figure}[h!]
  \centering
  \begin{subfigure}[t]{.5\linewidth}
    \centering
    \includegraphics[width=\linewidth]{processed_input_images_Pr_Po_Im_128_px}
    \caption{Scenario \texttt{Pr\_Po\_Im}}
    \label{pr_po_im_64}
  \end{subfigure}%
  \begin{subfigure}[t]{.5\linewidth}
    \centering
    \includegraphics[width=\linewidth]{processed_input_images_Pr_Im_128_px}
    \caption{Scenario \texttt{Pr\_Im}}
    \label{pr_im_64}
  \end{subfigure}%

  \begin{subfigure}[t]{.5\linewidth}
    \centering
    \includegraphics[width=\linewidth]{processed_input_images_PrPo_Im_128_px}
    \caption{Scenario \texttt{PrPo\_Im}}
    \label{prpo_im_64}
  \end{subfigure}%
  \begin{subfigure}[t]{.5\linewidth}
    \centering
    \includegraphics[width=\linewidth]{processed_input_images_Pr_PoIm_128_px}
    \caption{Scenario \texttt{Pr\_PoIm}}
    \label{pr_poim_64}
  \end{subfigure}%

  \caption{Random examples of the processed training images under each classification scenario; 64 pixels.}
  \label{fig:processed_images}
\end{figure}

@@

The various hyperparameters in the model are summarized in \autoref{tab:hyp}.
\begin{table}[]\small
\rowcolors{1}{white}{gray!35}
    \centering
    \begin{tabular}{l c c c c c c c}\toprule
    \bf Layer & \bf Layer \# & \bf No.\ Filters & \bf Kernel Size & \bf Strides &
     &  & \bf Activation \\\midrule
    Convolutional  & 1         & 64                & $k^*$   & 2           &&& ReLu \\
    Max. Pooling   & 1         &                   & 2               &&&             &      \\
    Convolutional  & 2         & 128               & 3   & 1           &&& ReLu \\
    Convolutional  & 3         & 128               & 3   & 1           &&& ReLu \\
    Max. Pooling   & 2         &                   & 2               &&&             &      \\    
    Convolutional  & 4         & 256               & 3   & 1           &&& ReLu \\
    Convolutional  & 5         & 256               & 3   & 1           &&& ReLu \\
    Max. Pooling   & 3         &                   & 2               &             &&&      \\\midrule
    Flatten & & & & &&& \\\midrule
     &  &   &  & &
     \bf Rate & \bf No.\ Units & \\\midrule
    Dense           & 1  & & &  & & $u_1^*$ & $a_1^*$ \\
    Dropout         & 1  & & &  & $r_1^*$ & &  \\
    Dense           & 2  & & &  & & $u_2^*$ & $a_2^*$ \\
    Dropout         & 2  & & &  & $r_2^*$ & & \\ \bottomrule
    \end{tabular}
    \caption{Summary of the convolutional neural network hyperparameters. 
    Those indicated by an asterisked symbol are optimized using a guided search.}
    \label{tab:hyp}
\end{table}

\subsection{Hyperparameter optimization}
We optimized eight of the CNN hyperparameters using Hyperband \cite{li2018hyperband}, an efficient guided grid-search algorithm.
Twelve searches were performed for each classification scenario and image resolution combination.
Each search was conducted using 90 trials of unique hyperparameter combinations. The specified range of each parameter along with the search results are shown in \autoref{tab:opt-hyp}.

@@

\begin{tabular}{@{}lllrrr@{}}

@@

                          & 1st dense activation, $a_1$  & \{ReLu, $\tanh$\}         & relu                    & tanh                    & relu                    \\
                          & 2nd dense activation, $a_2$  & \{ReLu, $\tanh$\}         & relu                    & relu                    & relu                    \\

@@

                          & 1st dense activation, $a_1$  & \{ReLu, $\tanh$\}         & relu                    & relu                    & relu                    \\
                          & 2nd dense activation, $a_2$  & \{ReLu, $\tanh$\}         & relu                    & relu                    & relu                    \\

@@

                          & 1st dense activation, $a_1$  & \{ReLu, $\tanh$\}         & relu                    & relu                    & relu                    \\
                          & 2nd dense activation, $a_2$  & \{ReLu, $\tanh$\}         & tanh                    & relu                    & tanh                    \\

@@

                          & 1st dense activation, $a_1$  & \{ReLu, $\tanh$\}         & relu                    & tanh                    & relu                    \\
                          & 2nd dense activation, $a_2$  & \{ReLu, $\tanh$\}         & tanh                    & relu                    & relu                    \\

@@

% \multirow{8}{*}{PrPo\_Im}   & activation\_1  & relu              & tanh              & relu              \\
%                             & activation\_2  & relu              & relu              & relu              \\

@@

% \multirow{8}{*}{Pr\_Im}     & activation\_1  & relu              & relu              & relu              \\
%                             & activation\_2  & relu              & relu              & relu              \\

@@

% \multirow{8}{*}{Pr\_PoIm}   & activation\_1  & relu              & relu              & relu              \\
%                             & activation\_2  & tanh              & relu              & tanh              \\

@@

% \multirow{8}{*}{Pr\_Po\_Im} & activation\_1  & relu              & tanh              & relu              \\\midrule
%                             & activation\_2  & tanh              & relu              & relu              \\

@@

\subsection{Model visualization and inference}
Model explainability is increasingly critical in the application of CNNs to an ever-expanding set of problems.

\begin{figure}[ht]
    \centering

    \begin{subfigure}[t]{.45\linewidth}
    \centering
    \includegraphics[width=\linewidth]{opt-gradcam-Pr_Im-128-px-4-images.png}
    \caption{Scenario \texttt{Pr\_Im}}
    \label{prpo_im_cm}
    \end{subfigure}%   
    \begin{subfigure}[t]{.45\linewidth}
    \centering
    \includegraphics[width=\linewidth]{opt-gradcam-PrPo_Im-128-px-4-images.png}
    \caption{Scenario \texttt{PrPo\_Im}}
    \label{prpo_im_cm}
    \end{subfigure}%

    \begin{subfigure}[t]{.45\linewidth}
    \centering
    \includegraphics[width=\linewidth]{opt-gradcam-Pr_PoIm-128-px-4-images.png}
    \caption{Scenario \texttt{Pr\_PoIm}}
    \label{prpo_im_cm}
    \end{subfigure}%
    \begin{subfigure}[t]{.45\linewidth}
    \centering
    \includegraphics[width=\linewidth]{opt-gradcam-Pr_Po_Im-128-px-4-images.png}
    \caption{Scenario \texttt{Pr\_Po\_Im}}
    \label{prpo_im_cm}
    \end{subfigure}%

    \caption{Gradient-weighted class activation maps (Grad-CAMs) for randomly selected validation images in each class. (Resolution: 128px)}
    \label{fig:my_label}
\end{figure}

@@

\section{Conclusion}

>>>>

\begin{figure}[h!]
  \centering
  \begin{subfigure}[t]{.75\linewidth}
    \centering
    \includegraphics[width=.8\linewidth, trim={0 3cm 0 3cm}, clip]{processed_input_images_Pr_Po_Im_128_px}
    \caption{Scenario \texttt{Pr\_Po\_Im}}
    \label{pr_po_im_64}
  \end{subfigure}%

  \begin{subfigure}[t]{.5\linewidth}
    \centering
    \includegraphics[width=\linewidth, trim={0 3cm 0 2cm}, clip]{processed_input_images_Pr_Im_128_px}
    \caption{Scenario \texttt{Pr\_Im}}
    \label{pr_im_64}
  \end{subfigure}%

  \begin{subfigure}[t]{.5\linewidth}
    \centering
    \includegraphics[width=\linewidth, trim={0 3cm 0 2cm}, clip]{processed_input_images_PrPo_Im_128_px}
    \caption{Scenario \texttt{PrPo\_Im}}
    \label{prpo_im_64}
  \end{subfigure}%
  \begin{subfigure}[t]{.5\linewidth}
    \centering
    \includegraphics[width=\linewidth, trim={0 3cm 0 2cm}, clip]{processed_input_images_Pr_PoIm_128_px}
    \caption{Scenario \texttt{Pr\_PoIm}}
    \label{pr_poim_64}
  \end{subfigure}%

  \caption{Random examples of the processed training images under each classification scenario; 128 pixels.}
  \label{fig:processed_images}
\end{figure}

@@

The various hyperparameters in the model are summarized in \autoref{tab:hyp}.
\begin{table}[]\small
\rowcolors{1}{white}{gray!35}
    \centering
    \begin{tabular}{l c c c c c c c}\toprule
    \bf Layer & \bf Layer \# & \bf No.\ Filters & \bf Kernel Size & \bf Strides &
     &  & \bf Activation \\\midrule
    Convolutional  & 1         & 64                & $k^*$   & 2           &&& ReLU \\
    Max. Pooling   & 1         &                   & 2               &&&             &      \\
    Convolutional  & 2         & 128               & 3   & 1           &&& ReLU \\
    Convolutional  & 3         & 128               & 3   & 1           &&& ReLU \\
    Max. Pooling   & 2         &                   & 2               &&&             &      \\    
    Convolutional  & 4         & 256               & 3   & 1           &&& ReLU \\
    Convolutional  & 5         & 256               & 3   & 1           &&& ReLU \\
    Max. Pooling   & 3         &                   & 2               &             &&&      \\\midrule
    Flatten & & & & &&& \\\midrule
     &  &   &  & &
     \bf Rate & \bf No.\ Units & \\\midrule
    Dense           & 1  & & &  & & $u_1^*$ & $a_1^*$ \\
    Dropout         & 1  & & &  & $r_1^*$ & &  \\
    Dense           & 2  & & &  & & $u_2^*$ & $a_2^*$ \\
    Dropout         & 2  & & &  & $r_2^*$ & & \\ \bottomrule
    \end{tabular}
    \caption{Summary of the convolutional neural network hyperparameters. 
    Those indicated by an asterisked symbol are optimized using a guided search.}
    \label{tab:hyp}
\end{table}

\subsection{Hyperparameter optimization}
We optimized eight of the CNN hyperparameters using Hyperband \cite{li2018hyperband}, an efficient guided grid-search algorithm.
Twelve searches were performed for each classification scenario and image resolution combination.
Each search was conducted using 90 trials of unique hyperparameter combinations. The specified range of each parameter along with the search results are shown in \autoref{tab:opt-hyp}. For the kernel size in the first convolutional layer, we allowed for a choice between a $5\times 5$ and a $7\times 7$ kernel. The activation function in both dense layers was specified as a choice between the rectified linear unit (ReLU) function and the hyperbolic tangent ($\tanh$). The ReLU was introduced to address the so-called "vanishing gradient" problem and has been shown to improve performance in CNNs \cite{glorot2011deep}. Nevertheless, the $\tanh$ function remains a viable option, as well. The dropout rates were allowed to range from 0 to 5 in steps of 0.05, while the number of neurons or units in each dense layer varied from 32 to 512 in steps of 32. Finally, we uniformly sampled learning rates $\la$ for the optimizer in the $\log_{10}$ space of $[10^{-4}, 10^{-2}]$.

@@

\begin{tabular}{@{}lllccc@{}}

@@

                          & 1st dense activation, $a_1$  & \{ReLU, $\tanh$\}         & ReLU                    & tanh                    & ReLU                    \\
                          & 2nd dense activation, $a_2$  & \{ReLU, $\tanh$\}         & ReLU                    & ReLU                    & ReLU                    \\

@@

                          & 1st dense activation, $a_1$  & \{ReLU, $\tanh$\}         & ReLU                    & ReLU                    & ReLU                    \\
                          & 2nd dense activation, $a_2$  & \{ReLU, $\tanh$\}         & ReLU                    & ReLU                    & ReLU                    \\

@@

                          & 1st dense activation, $a_1$  & \{ReLU, $\tanh$\}         & ReLU                    & ReLU                    & ReLU                    \\
                          & 2nd dense activation, $a_2$  & \{ReLU, $\tanh$\}         & tanh                    & ReLU                    & tanh                    \\

@@

                          & 1st dense activation, $a_1$  & \{ReLU, $\tanh$\}         & ReLU                    & tanh                    & ReLU                    \\
                          & 2nd dense activation, $a_2$  & \{ReLU, $\tanh$\}         & tanh                    & ReLU                    & ReLU                    \\

@@

% \multirow{8}{*}{PrPo\_Im}   & activation\_1  & ReLU              & tanh              & ReLU              \\
%                             & activation\_2  & ReLU              & ReLU              & ReLU              \\

@@

% \multirow{8}{*}{Pr\_Im}     & activation\_1  & ReLU              & ReLU              & ReLU              \\
%                             & activation\_2  & ReLU              & ReLU              & ReLU              \\

@@

% \multirow{8}{*}{Pr\_PoIm}   & activation\_1  & ReLU              & ReLU              & ReLU              \\
%                             & activation\_2  & tanh              & ReLU              & tanh              \\

@@

% \multirow{8}{*}{Pr\_Po\_Im} & activation\_1  & ReLU              & tanh              & ReLU              \\\midrule
%                             & activation\_2  & tanh              & ReLU              & ReLU              \\

@@

% \subsection{Model visualization and inference}
% Model explainability is increasingly critical in the application of CNNs to an ever-expanding set of problems.

% \begin{figure}[ht]
%     \centering

%     \begin{subfigure}[t]{.45\linewidth}
%     \centering
%     \includegraphics[width=\linewidth]{opt-gradcam-Pr_Im-128-px-4-images.png}
%     \caption{Scenario \texttt{Pr\_Im}}
%     \label{prpo_im_cm}
%     \end{subfigure}%   
%     \begin{subfigure}[t]{.45\linewidth}
%     \centering
%     \includegraphics[width=\linewidth]{opt-gradcam-PrPo_Im-128-px-4-images.png}
%     \caption{Scenario \texttt{PrPo\_Im}}
%     \label{prpo_im_cm}
%     \end{subfigure}%

%     \begin{subfigure}[t]{.45\linewidth}
%     \centering
%     \includegraphics[width=\linewidth]{opt-gradcam-Pr_PoIm-128-px-4-images.png}
%     \caption{Scenario \texttt{Pr\_PoIm}}
%     \label{prpo_im_cm}
%     \end{subfigure}%
%     \begin{subfigure}[t]{.45\linewidth}
%     \centering
%     \includegraphics[width=\linewidth]{opt-gradcam-Pr_Po_Im-128-px-4-images.png}
%     \caption{Scenario \texttt{Pr\_Po\_Im}}
%     \label{prpo_im_cm}
%     \end{subfigure}%

%     \caption{Gradient-weighted class activation maps (Grad-CAMs) for randomly selected validation images in each class. (Resolution: 128px)}
%     \label{fig:my_label}
% \end{figure}

@@

\section{Conclusion}
We have demonstrated the efficacy of an artificial intelligence framework for predicting tree failure likelihood (Probable, Possible and Improbable) with respect to utility infrastructure. Specifically, we developed a convolutional neural network with state-of-the-art configurations.
We applied data augmentation and pre-processing strategies to increase the size of our dataset by a factor of 5, thus generating 2525 images.
From an initial resolution of $4032 \times 3024$ pixels, we created three sets of image resolutions: $64 \times 64$, $128 \times 128$ and $224 \times 224$.
We then optimized eight of our CNN hyperparameters for 12  classification scenario and image resolution combinations. 
We conducted ten model training trials in each of the 12 cases, using a 20\% validation set to measure model performance.
In these trials, the average accuracy of classification was $\geq 0.94$, with a maximum standard deviation of 0.7.
There was no statistically significant difference between the performance across image resolutions.

We further conducted more detailed classification performance analyses over single training instances for the 128-pixel case.
Our results indicate that the CNN performed best at recalling Probable vs. Improbable cases.
The Possible category appeared to be a confounding case.
This was not unexpected, given the uncertainty and subjectivity in assessing these trees to begin with.
The CNN, however, performed better at distinguishing between Possible/Improbable and Probable, than between Probable/Possible and Improbable.
This might be an indicator that the Possible cases are more likely identified as Improbable.
When all three failure-likelihood categories were predicted individually, the Possible case had the lowest recall score.

Nevertheless, given the relatively small input dataset of original images, these preliminary results are extremely promising for future improvements.
First, we can train better models with more data.
We also plan to rigorously quantify the uncertainty in ground truth category assignments by incorporating predictions from multiple experts on the same images.
In order to better understand how the CNN is classifying each image, we will conduct extensive visual inference in further work, for instance, using the gradient-weighted class activation mapping approach \cite{zeiler2014visualizing}.
